---
title: "Inflation 2025 Forecasting"
author: "Car3lo"
date: "2025-05-18"
output: github_document
---

# Forecasting 2025 PH Inflation Rates

### Load Libraries and Dataset
```{r}
library(tidyverse)
library(forecast)
library(fpp2)
library(aTSA)
library(FinTS)
library(lmtest)
setwd("D:/New Documents/UPSE_4-2/portfolio_temp/[132] Remittance and Inflation Forecasts/Inflation")

```

### Creating Objects
```{r}
inf <- read.csv("inf_ph.csv")
inf_ph <- ts(inf[,2], frequency = 12, start = c(2018,1), end = c(2025,4))
inf_ph
```

### Plottingg
```{r}
autoplot(inf_ph)+
  xlab("Month")+
  ylab("Inflation Rate (%)")+
  ggtitle("Philippine Monthly Headline Inflation Jan 2028 - Apr 2025")
```

- Inflation had uptrend-peak-downtrend patterns within 2-3 years.
- This suggests cyclic behavior from it.
- There may also be a hint of weak seasonality.
- They also peak and bottom near 7.5% and 2.5% respectively implying corrective policy changes

### Checking for Stationarity
```{r}
ggtsdisplay(inf_ph)
# Augmented Dickey-Fuller (ADF) test
floor(12*(length(inf_ph)/100)^(1/4))+1 
adf.test(inf_ph, nlag=12) 
```

- We will use Type 3 since from the autoplot, the inflation series has cyclic behavior
- HO: Series has unit-root; non-stationary
- HA: Series is stationary

- Decision: We reject HO if all p-values of lags up to 12 in the ADF test's selected Type is < 0.05
- Conclusion: All p-values are above 0.05, so we do not reject HO. The inflation series is non-stationary.


### First-Difference Transformation and Checking for Stationarity
```{r}
dinf_ph <- diff(inf_ph,1)
autoplot(dinf_ph)
adf.test(dinf_ph, nlag=12) 
```

- We will use Type 1 since from the autoplot, the inflation series seem to have no trend/drift
- HO: Series has unit-root; non-stationary
- HA: Series is stationary

- Conclusion: All p-values are below 0.05 (around 0.01) -- we reject Ho. It is now stationary.


### ARIMA MODELLING AND RESIDUAL DIAGNOSTICS
```{r}
inf_ph_auto <- auto.arima(inf_ph)
checkresiduals(inf_ph_auto)
Box.test(residuals(inf_ph_auto), type="Ljung-Box")
shapiro.test(residuals(inf_ph_auto))
ArchTest(residuals(inf_ph_auto))
```

- We fail to reject the Null Hypothesis that the residuals are white noise. This is consistent with the ACF graph.
- We fail to reject the Null Hypothesis that the residuals are normally distributed. This is consistent with the histogram of its residuals.
- We fail to reject the Null Hypothesis that the residuals are homoskedastic.


### The Model
- Model suggests ARIMA(3,0,1)(2,0,0)[12] with non-zero mean model.
- ARIMA(3,0,1) non-seasonal part:
- - 3 AR Terms (current value depends on past 3 observations)
- - 0 differencing
- - 1 MA terms (accounts for single lagged error term)

- ARIMA (2,0,0)[12] seasonal part: 
- - 2 Seasonal AR Terms (12 months ago with two-lag effect)
- - 0 seasonal differencing
- - 0 seasonal MA terms
- - [12] period of seasonal pattern

- Non-zero includes constant or trend component.

#### Testing the model further
```{r}
autoplot(inf_ph_auto)
coeftest(inf_ph_auto)
```

- From the z test of coefficients, all are significant (not less than 0) at at least 5% significance level 
- EXCEPT for AR3 -- the third lag of AR terms do not contribute significantly to the model


### Plotting Fitted vs Actual
```{r}
inf_ph_auto_fit <- fitted(inf_ph_auto)
autoplot(inf_ph, series="Actual") +
  autolayer(inf_ph_auto_fit, series="Fitted") +
  ggtitle("Actual vs Fitted Inflation") +
  xlab("Month") + ylab("Inflation Rate (%)")
```

### Trying simpler models
```{r}
# AR(4)
inf_ph_AR4 <- arima(dinf_ph, order = c(4,0,0))

Box.test(residuals(inf_ph_AR4), type = "Ljung-Box")
shapiro.test(residuals(inf_ph_AR4))
ArchTest(residuals(inf_ph_AR4))
autoplot(inf_ph_AR4)
coeftest(inf_ph_AR4)
```

```{r}
# MA(2)
inf_ph_MA2 <- arima(dinf_ph, order = c(0, 0, 2))

Box.test(residuals(inf_ph_MA2), type = "Ljung-Box")
shapiro.test(residuals(inf_ph_MA2))
ArchTest(residuals(inf_ph_MA2))
autoplot(inf_ph_MA2)
coeftest(inf_ph_MA2)
```

```{r}
# ARMA(1,1)
inf_ph_ARMA11 <- arima(dinf_ph, order = c(1,0,1))

Box.test(residuals(inf_ph_ARMA11), type = "Ljung-Box")
shapiro.test(residuals(inf_ph_ARMA11))
ArchTest(residuals(inf_ph_ARMA11))
autoplot(inf_ph_ARMA11)
coeftest(inf_ph_ARMA11)
```

### Trying ARMA(1,1)+SAR(1)
```{r}
inf_ph_ARMA11_SAR1 <- arima(dinf_ph, order = c(1, 0, 1), seasonal = list(order = c(1, 0, 0), period = 12))
Box.test(residuals(inf_ph_ARMA11_SAR1), type = "Ljung-Box")
shapiro.test(residuals(inf_ph_ARMA11_SAR1)) 
ArchTest(residuals(inf_ph_ARMA11_SAR1))
autoplot(inf_ph_ARMA11_SAR1)
coeftest(inf_ph_ARMA11_SAR1)
```

#### **FOR ARMA(1,1)+SAR(1)
- At 5% Significance we fail to reject HO (The residuals are white noise)
- At 5% Significance we fail to reject HO (The residuals are normally distributed)
- At 5% significance we fail to reject HO (The residuals are homoskedastic)
- All inverse AR and MA roots are within their unit circles.
- AR(1), MA(1), and SAR(1) are all significant at atleast 5% significance level.
- Past values, short-term shocks, and last year's inflation affect current inflation in the series.


### Other Models and their Coefficients
![Table of Equations](https://lh3.googleusercontent.com/pw/AP1GczNy8Xt_28pr7u-lURn7w_235gtxjKz7N-nsURAJY1X3me5IeWgnH9pqI76NjxDJLxcZODSPNqoC1cj7qMeW02z_a0qNMjj2EKfH9QeupaPa-R6g_DA12jA4T248aNPBnMhc9P6_THqpB-TgP9_t0uyW6Xn0qJqDMfhHIw4aCq1wowL92wVAfH-OH6tpBJPq5UVZh1IMF8sQ3JehRk1Iy2-IyyRGNWurOGzL3ZQwr54NzwkeoN79DCMGh4Itpj2K0V3f5DPavWgHif2PHtX7pQQyB-XQPZpR2jFk2vbNXZCo673rvtNWzCi5zegjLAjnV88F5FBom_vnFcxo_v6n5RDso-6CgEilnznHCG4REjqiU1XFGQfRbhxqQ2BEpPyD416XYP73zcx4-wphmjZTO1S8mLkKdV3erwH2LRZhKWeBYvzwFOa1YG6QXk5_7uxMCWzATyuF1F3_vmPeyo0pcpVVJy0DLueNfoISueFeTMckiPK7Djs0BZCbZi_TzdReMQ4m5hY_DCWiBn17ufKtDOF9fkgcEjeMug2Nzc_5krpvEJHIZbnlMjFkygRsiBejm8k00g1s9bNQF72Z14lvkidJf6Mqg6Du-2zmFDttAGCpiabwdjThnzGitBYpAekUeh0pvLfWAkH4uclnZdns7txxCKnRshTJ1XMfRWzuytH0ubG28aeIZ3fB5w8n9EwxU3jd2XOI9svEbvitWGM1Gv3rLNO3-BuMS8vUJhhx76uVFGpalI3ljgw5NK1oxXgjYluYSqRS-vmxdD15YuFDhuusTPgHLph5JBHFaHfIAWMTUxgkzjTtiKBHRsOpE0cTLoTzMkfUwcbvy0LMPiOVw7q5W15ANQU3hJMTvFbRIqoTXdvs5vsnXXkDg1CgFf1Cd_VeBj-_dzIgCqBgppIiPsj_9fUeY90xHlTxxsDbk-DdPFK1e7crgriT3YOYlo_oasbQB0ovAFF-Oa8Rd4BKVOMoZyKoYwTs5SUTMTUw4WF740B16iy07EAOlpNnvrr0taWwWGOntUIgWm-_ATaO_Rzex7IJBqdtCrOiLZApsNWYCETvarnaDwoHVcPTc2dCrlXRb57TbtEYLiab4oA83MMIn1ZDheQAGzCO3ZDME5Rk8zUR5RwQDMJlgNGafAE6VEu-dkZPStv-9-r_VOLsR5MUAJ2bow_-j-YVNkO-=w966-h326-no?authuser=0)


### Plotting the Models
```{r}
autoplot(dinf_ph, series = "Actual") +
  autolayer(fitted(inf_ph_AR4), series = "AR(4)") +
  autolayer(fitted(inf_ph_MA2), series = "MA(2)") +
  autolayer(fitted(inf_ph_ARMA11), series = "ARMA(1,1)") +
  autolayer(fitted(inf_ph_ARMA11_SAR1), series = "ARMA(1,1) + SAR(1)") +
  xlab("Month") + 
  ylab("Inflation Rate (%)") +
  ggtitle("Actual vs Fitted Inflation Models (with differencing)")
```
- All models seem close to fitting actual differenced inflation values.
- The closest one seems to be ARMA(1,1) + SAR(1) Model

### Comparing Accuracy
```{r}
# RMSE MAE MAPE
accuracy(inf_ph_auto)
accuracy(inf_ph_AR4)
accuracy(inf_ph_MA2)
accuracy(inf_ph_ARMA11)
accuracy(inf_ph_ARMA11_SAR1)

# AIC, AICc, BIC
AIC(inf_ph_auto)
BIC(inf_ph_auto)
inf_ph_auto$aicc

Arima(dinf_ph, order = c(4, 0, 0), include.mean = TRUE)
Arima(dinf_ph, order = c(0, 0, 2), include.mean = TRUE)
Arima(dinf_ph, order = c(1, 0, 1), include.mean = TRUE)
Arima(dinf_ph, order = c(1, 0, 1), seasonal = c(1, 0, 0), include.mean = TRUE)
```

- Although lowest RMSE and MAE belong to autoarima, ARMA(1,1)+SAR(1) has lowest AIC, AICc, BIC.
- Plus, the RMSE and MAE are pretty close anyway.
- So we will choose ARMA(1,1)+SAR(1) as our "best fitting" model.

![Accuracy](https://lh3.googleusercontent.com/pw/AP1GczOCPLNycpMJFZHTJUpkRtJfEjbPqO86yns4TTEgI9J6AFV7qLbt4L11zkBqhC9ZIRzrqb7QvQjPXCq2ar7ZeVpNUumKI_FFG2Tm8q5JG1UUBoNKkA-M2Oci0qa451mVgcwqgjD-MkXnR2aUCcLB_k2w=w1817-h703-s-no-gm?authuser=0)


### Forecasting and Plotting First Difference
```{r}
inf_ph_ARMA11_SAR1_forecast <- forecast::forecast(inf_ph_ARMA11_SAR1, h = 8, level = 95) #conflicts with aTSA
autoplot(inf_ph_ARMA11_SAR1_forecast) +
  autolayer(dinf_ph, series = "Actual") +
  xlab("Month")+
  ylab("Inflation Rate")+
  ggtitle("Inflation Forecast Percentage Point Difference, May-Dec 2025")
```


### Plotting Headline Inflation
```{r}
# get first differences from object
diff_forecast <- inf_ph_ARMA11_SAR1_forecast$mean
diff_lower <- inf_ph_ARMA11_SAR1_forecast$lower[,1]
diff_upper <- inf_ph_ARMA11_SAR1_forecast$upper[,1]

# convert to headline inflation with april 2025 value (1.4)
headline_forecast <- 1.4 + cumsum(diff_forecast)
headline_lower <- 1.4 + cumsum(diff_lower)
headline_upper <- 1.4 + cumsum(diff_upper)

# make it timeseries
ts_forecast <- ts(headline_forecast, start = c(2025, 5), frequency = 12)
ts_lower <- ts(headline_lower, start = c(2025, 5), frequency = 12)
ts_upper <- ts(headline_upper, start = c(2025, 5), frequency = 12)

# object with class forecast
fc <- list(mean = ts_forecast,
           lower = matrix(ts_lower, ncol=1),
           upper = matrix(ts_upper, ncol=1),
           level = 95,
           x = inf_ph)  
class(fc) <- "forecast"

# plotting time
autoplot(fc) + 
  autolayer(inf_ph, series = "Actual") +
  xlab("Month") +
  ylab("Headline Inflation Rate (%)") +
  ggtitle("Forecasted Headline Inflation (May-Dec 2025)")
```

[Inflation Foercasts](https://lh3.googleusercontent.com/pw/AP1GczN_sOjFLAtQOZiikSnoI0-jW1S1wWbmm3wJELipiVUAUJnjMDZPkFyPq7CwJrz5IDVVXHQ9aN0_ovyc1llLpDB9bSURZN1_98vvUf0uI6wfXg4UzM_dNCnVg7njpQHWrhid5RoNz0olKQ0L5k1ucXRo=w1156-h832-s-no-gm?authuser=0)

### Key Results
- The 95% Interval Forecast for May 2025 is [0.4760, 2.1029]
- Average inflation rate for the rest of 2025 (forecasted) is 1.561275 (Not within BSP's Target)

